Reanalysis Step 9: Alternative Propensity Score Specifications

### 2025-09-05 reanalysis-9-alternative_specifications.qmd
reanalysis streategy pattern

@@

  # Check cache
  if (use_cache && file.exists(cache_file)) {
    cat(paste0("  Loading cached results for ", method_name, " (imp ", imputation_id, ")\n"))
    return(readRDS(cache_file))
  }

  cat(paste0("  Estimating ", method_name, " for imputation ", imputation_id, "\n"))

  result <- tryCatch({
    if (method_params$package == "matchit") {
      # MatchIt methods
      args <- list(
        formula = ps_formula,
        data = data,
        method = method_params$method,
        estimand = method_params$estimand
      )

      # Add method-specific parameters
      if (!is.null(method_params$distance)) args$distance <- method_params$distance
      if (!is.null(method_params$replace)) args$replace <- method_params$replace
      if (!is.null(method_params$subclass)) args$subclass <- method_params$subclass
      if (!is.null(method_params$distance.options)) args$distance.options <- method_params$distance.options

      do.call(matchit, args)

    } else if (method_params$package == "weightit") {
      # WeightIt methods
      args <- list(
        formula = ps_formula,
        data = data,
        method = method_params$method,
        estimand = method_params$estimand
      )

      # Add method-specific parameters
      if (!is.null(method_params$over)) args$over <- method_params$over

      do.call(weightit, args)

    } else {
      stop(paste("Unknown package:", method_params$package))
    }
  }, error = function(e) {
    cat(paste0("  ERROR in ", method_name, " (imp ", imputation_id, "): ", e$message, "\n"))
    return(NULL)
  })

  # Save to cache
  if (!is.null(result) && use_cache) {
    saveRDS(result, cache_file)
  }

  return(result)
}

# Function to process all imputations for a single method
process_method_all_imputations <- function(method_params, imputed_datasets, ps_formula, 
                                          cache_dir, use_cache = TRUE, n_cores = 1) {
  n_imp <- length(imputed_datasets)
  method_name <- method_params$name

  log_message(paste0("Processing method: ", method_name, " across ", n_imp, " imputations"))

  if (n_cores > 1 && .Platform$OS.type == "unix") {
    # Parallel processing for Unix/Mac
    results <- mclapply(
      1:n_imp,
      function(i) process_ps_method(i, imputed_datasets[[i]], method_params, ps_formula, cache_dir, use_cache),
      mc.cores = n_cores
    )
  } else if (n_cores > 1 && .Platform$OS.type != "unix") {
    # Parallel processing for Windows
    cl <- makeCluster(n_cores)
    clusterExport(cl, c("process_ps_method", "ps_formula", "cache_dir", "use_cache", "method_params"),
                  envir = environment())
    clusterEvalQ(cl, {
      library(MatchIt)
      library(WeightIt)
    })

    results <- parLapply(cl, 1:n_imp, function(i) {
      process_ps_method(i, imputed_datasets[[i]], method_params, ps_formula, cache_dir, use_cache)
    })

    stopCluster(cl)
  } else {
    # Sequential processing
    results <- lapply(
      1:n_imp,
      function(i) process_ps_method(i, imputed_datasets[[i]], method_params, ps_formula, cache_dir, use_cache)
    )
  }

  return(results)
}
```

## Run Propensity Score Methods

```{r run-ps-methods}
# Process only the specified number of imputations
imputed_datasets_subset <- imputed_datasets[1:min(max_imputations_to_process, length(imputed_datasets))]

# Initialize results storage
all_ps_results <- list()

# Process each method sequentially (to manage memory)
for (method_params in ps_methods) {
  method_name <- method_params$name
  log_message(paste0("\n", paste(rep("=", 50), collapse = "")))
  log_message(paste0("Starting method: ", method_name))

  # Process all imputations for this method
  method_results <- process_method_all_imputations(
    method_params = method_params,
    imputed_datasets = imputed_datasets_subset,
    ps_formula = ps_formula,
    cache_dir = cache_dir,
    use_cache = use_cache,
    n_cores = max_cores_for_ps
  )

  all_ps_results[[method_name]] <- method_results

  # Clean up memory after each method
  gc()

  log_message(paste0("Completed method: ", method_name))
}

log_message("All PS methods completed")
```

## Load twang_gbm Results from Reanalysis-4

```{r load-twang}
# Load pre-computed twang_gbm results
twang_cache_dir <- file.path(reanalysis_data_dir, "ps_cache")
twang_results <- list()

for (i in 1:max_imputations_to_process) {
  twang_file <- file.path(twang_cache_dir, paste0("ps_imp_", i, ".rds"))
  if (file.exists(twang_file)) {
    twang_results[[i]] <- readRDS(twang_file)
    log_message(paste0("Loaded twang_gbm results for imputation ", i))
  } else {
    log_message(paste0("WARNING: twang_gbm results not found for imputation ", i))
    twang_results[[i]] <- NULL
  }
}

# Add to all results
all_ps_results[["twang_gbm"]] <- twang_results

log_message(paste0("Loaded twang_gbm results for ", sum(!sapply(twang_results, is.null)), " imputations"))
```

## Outcome Analysis

```{r outcome-analysis}
# Get outcome formula from functions.R
outcome_formulas <- get_analysis_formulas_reanalysis()
outcome_formula <- outcome_formulas$full

log_message("Starting outcome analysis for all methods")

# Function to fit outcome model for a single method and imputation
fit_outcome_model <- function(ps_result, data, formula, method_name) {
  if (is.null(ps_result)) {
    return(NULL)
  }

  tryCatch({
    if (method_name == "twang_gbm") {
      # twang weights
      weights <- ps_result$w[, "es.mean.ATT"]
      data$weights <- weights
      design <- svydesign(ids = ~e_mrn_deidentified, weights = ~weights, data = data)

    } else if (inherits(ps_result, "matchit")) {
      # MatchIt results
      if (grepl("subclass", method_name)) {
        # Subclassification
        matched_data <- match.data(ps_result, subclass = "subclass")
        design <- svydesign(ids = ~e_mrn_deidentified, data = matched_data, weights = ~weights)
        design <- postStratify(design, ~subclass, xtabs(~subclass, data = matched_data))
      } else {
        # Matching
        matched_data <- get_matches(ps_result)
        design <- svydesign(ids = ~e_mrn_deidentified, data = matched_data, weights = ~1)
      }

    } else if (inherits(ps_result, "weightit")) {
      # WeightIt results
      weights <- ps_result$weights
      data$weights <- weights
      design <- svydesign(ids = ~e_mrn_deidentified, weights = ~weights, data = data)

    } else {
      stop("Unknown result type")
    }

    # Fit the model
    model <- svyglm(formula, design = design, family = quasibinomial())

    # Extract coefficient for lapse
    coef_lapse <- coef(summary(model))["ever_lapse_binary", ]

    return(list(
      estimate = coef_lapse["Estimate"],
      se = coef_lapse["Std. Error"],
      model = model
    ))

  }, error = function(e) {
    cat(paste0("  Error fitting model for ", method_name, ": ", e$message, "\n"))
    return(NULL)
  })
}

# Process outcome models for all methods and imputations
outcome_results <- list()

for (method_name in names(all_ps_results)) {
  log_message(paste0("Fitting outcome models for ", method_name))

  method_outcomes <- list()
  ps_results <- all_ps_results[[method_name]]

  for (i in 1:length(ps_results)) {
    if (!is.null(ps_results[[i]])) {
      outcome <- fit_outcome_model(
        ps_result = ps_results[[i]],
        data = imputed_datasets_subset[[i]],
        formula = outcome_formula,
        method_name = method_name
      )
      method_outcomes[[i]] <- outcome
    }
  }

  outcome_results[[method_name]] <- method_outcomes
}

log_message("Outcome analysis completed")
```

## Pool Results Using Rubin's Rules

```{r pool-results}
# Function to pool results across imputations using Rubin's rules
pool_results <- function(method_outcomes) {
  # Filter out NULL results
  valid_outcomes <- method_outcomes[!sapply(method_outcomes, is.null)]

  if (length(valid_outcomes) == 0) {
    return(NULL)
  }

  # Extract estimates and SEs
  estimates <- sapply(valid_outcomes, function(x) x$estimate)
  ses <- sapply(valid_outcomes, function(x) x$se)

  # Remove any NA values
  valid_idx <- !is.na(estimates) & !is.na(ses)
  estimates <- estimates[valid_idx]
  ses <- ses[valid_idx]

  if (length(estimates) == 0) {
    return(NULL)
  }

  m <- length(estimates)  # Number of imputations

  # Pooled estimate (average of estimates)
  pooled_estimate <- mean(estimates)

  # Within-imputation variance (average of squared SEs)
  within_var <- mean(ses^2)

  # Between-imputation variance
  between_var <- var(estimates)

  # Total variance using Rubin's rules
  total_var <- within_var + between_var + (between_var / m)
  pooled_se <- sqrt(total_var)

  # Degrees of freedom (Barnard-Rubin adjustment)
  lambda <- (between_var + between_var/m) / total_var
  df_old <- (m - 1) / lambda^2

  # For large samples, use df_old
  df <- df_old

  # Calculate confidence intervals and p-value
  t_stat <- pooled_estimate / pooled_se
  p_value <- 2 * pt(abs(t_stat), df = df, lower.tail = FALSE)
  ci_lower <- pooled_estimate - qt(0.975, df) * pooled_se
  ci_upper <- pooled_estimate + qt(0.975, df) * pooled_se

  # Convert to odds ratio scale
  or <- exp(pooled_estimate)
  or_ci_lower <- exp(ci_lower)
  or_ci_upper <- exp(ci_upper)

  return(list(
    estimate = pooled_estimate,
    se = pooled_se,
    or = or,
    or_ci_lower = or_ci_lower,
    or_ci_upper = or_ci_upper,
    p_value = p_value,
    n_imputations = m,
    df = df
  ))
}

# Pool results for each method
pooled_results <- list()

for (method_name in names(outcome_results)) {
  pooled <- pool_results(outcome_results[[method_name]])
  if (!is.null(pooled)) {
    pooled$method <- method_name
    pooled_results[[method_name]] <- pooled
    log_message(paste0("Pooled results for ", method_name, ": OR = ", 
                      round(pooled$or, 3), " (", round(pooled$or_ci_lower, 3), 
                      ", ", round(pooled$or_ci_upper, 3), ")"))
  } else {
    log_message(paste0("WARNING: Could not pool results for ", method_name))
  }
}

# Save pooled results
saveRDS(pooled_results, file.path(reanalysis_data_dir, "alternative_ps_results_pooled.rds"))
log_message("Saved pooled results")
```

## Create Forest Plot

```{r forest-plot}
# Prepare data for forest plot
forest_data <- do.call(rbind, lapply(names(pooled_results), function(method) {
  res <- pooled_results[[method]]
  data.frame(
    method = method,
    or = res$or,
    lower = res$or_ci_lower,
    upper = res$or_ci_upper,
    p_value = res$p_value,
    stringsAsFactors = FALSE
  )
}))

# Order by odds ratio
forest_data <- forest_data[order(forest_data$or), ]

# Create more readable method names
forest_data$method_label <- gsub("_", " ", forest_data$method)
forest_data$method_label <- gsub("nearest ", "NN-", forest_data$method_label)
forest_data$method_label <- gsub("subclass ", "Subclass-", forest_data$method_label)
forest_data$method_label <- tools::toTitleCase(forest_data$method_label)

# Create forest plot using ggplot2
library(ggplot2)

forest_plot <- ggplot(forest_data, aes(x = or, y = reorder(method_label, or))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_point(size = 3, shape = 18) +
  scale_x_continuous(
    trans = "log",
    breaks = c(0.5, 0.75, 1, 1.5, 2, 3),
    labels = c("0.5", "0.75", "1.0", "1.5", "2.0", "3.0")
  ) +
  labs(
    x = "Odds Ratio (95% CI)",
    y = "Propensity Score Method",
    title = "Effect of Lapse on Vision Loss Across PS Methods",
    subtitle = paste0("Based on ", max_imputations_to_process, " imputations with Rubin's pooling")
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(hjust = 0)
  )

@@

# Save plot
ggsave(
  filename = file.path(reanalysis_data_dir, "forest_plot_ps_methods.png"),
  plot = forest_plot,
  width = 10,
  height = 8,
  dpi = 300
)

@@

## Summary Tables

```{r summary-tables}
# Create detailed summary table
summary_table <- do.call(rbind, lapply(names(pooled_results), function(method) {
  res <- pooled_results[[method]]
  data.frame(
    Method = method,
    `OR` = sprintf("%.3f", res$or),
    `95% CI` = sprintf("(%.3f, %.3f)", res$or_ci_lower, res$or_ci_upper),
    `P-value` = sprintf("%.4f", res$p_value),
    `N Imputations` = res$n_imputations,
    stringsAsFactors = FALSE
  )
}))

# Sort by OR
summary_table <- summary_table[order(as.numeric(gsub("([0-9.]+)", "\\1", summary_table$OR))), ]

kable(summary_table, 
      caption = "Pooled Effect Estimates Across Propensity Score Methods",
      col.names = c("Method", "Odds Ratio", "95% CI", "P-value", "N Imputations"),
      align = c("l", "r", "r", "r", "r"))

# Save to CSV
write.csv(summary_table, 
          file.path(reanalysis_data_dir, "ps_methods_comparison_table.csv"),
          row.names = FALSE)

log_message("Summary table saved")

# Create comparison with twang_gbm as reference
if ("twang_gbm" %in% names(pooled_results)) {
  ref_or <- pooled_results[["twang_gbm"]]$or

  comparison_table <- do.call(rbind, lapply(names(pooled_results), function(method) {
    res <- pooled_results[[method]]
    data.frame(
      Method = method,
      OR = res$or,
      `Relative to twang` = sprintf("%.1f%%", 100 * (res$or / ref_or - 1)),
      stringsAsFactors = FALSE
    )
  }))

  kable(comparison_table,
        caption = "Comparison of Methods Relative to twang_gbm",
        col.names = c("Method", "Odds Ratio", "% Difference from twang_gbm"),
        align = c("l", "r", "r"))
}
```

@@

This analysis compared `r length(pooled_results)` different propensity score methods across `r max_imputations_to_process` multiply imputed datasets. Results were pooled using Rubin's rules to account for between-imputation variability.

Key findings:
- Method with smallest effect: `r names(pooled_results)[which.min(sapply(pooled_results, function(x) x$or))]`
- Method with largest effect: `r names(pooled_results)[which.max(sapply(pooled_results, function(x) x$or))]`
- Range of odds ratios: `r sprintf("%.3f to %.3f", min(sapply(pooled_results, function(x) x$or)), max(sapply(pooled_results, function(x) x$or)))`

>>>>

**Analysis Configuration:**
- Mode: `r analysis_mode`
- Maximum imputations to process: `r config$n_imputations`
- Available imputations: `r n_imputations`
- Cache: `r ifelse(use_cache, "Enabled", "Disabled")`

@@

# Extract components
ps_results <- results$ps_results
outcome_results <- results$outcome_results
pooled_results <- results$pooled_results
forest_data <- results$forest_data
summary_table <- results$summary_table

# Save pooled results for later use
saveRDS(pooled_results, file.path(reanalysis_data_dir, "alternative_ps_results_pooled.rds"))
log_message("Saved pooled results")
```

## Forest Plot

```{r forest-plot}
#| fig.height: 10
#| fig.width: 8

# Create forest plot matching the style of results-summary.Rmd
# Add colors for specific methods
forest_data$color <- "black"
if ("twang_gbm" %in% forest_data$method) {
  forest_data$color[forest_data$method == "twang_gbm"] <- "blue"
}

# Add a naive/unadjusted estimate if we want to show it as reference
# (Would need to run this separately if desired)

# Create the plot
forest_plot <- ggplot(forest_data, aes(x = reorder(method_label, or), y = or)) +
  geom_pointrange(aes(ymin = lower, ymax = upper, color = color), size = 0.8) +
  geom_point(aes(color = color), size = 3) +
  geom_text(aes(label = sprintf("%.2f", or), color = color), 
            vjust = -1.2, hjust = 0.5, size = 4) +
  coord_flip() +
  scale_color_identity() +
  scale_y_continuous(breaks = seq(0.5, max(forest_data$upper) + 0.5, by = 0.1),
                     trans = "log",
                     labels = function(x) sprintf("%.1f", x)) +
  scale_x_discrete(expand = c(0.075, 0.075)) +
  theme_minimal(base_size = 14) +
  labs(
    x = element_blank(),
    y = "Odds Ratio",
    title = "Propensity Score Method Comparison with Multiple Imputation",
    subtitle = paste0("Based on ", config$n_imputations, " imputations with Rubin's pooling"),
    caption = paste0("\nAll methods use ATT (Average Treatment Effect in the Treated) estimand.\n",
                    "IPTW = Inverse probability of treatment weighting.\n",
                    ifelse("twang_gbm" %in% forest_data$method,
                          "twang GBM (shown in blue) is the primary method reported in the paper.",
                          ""))
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", margin = margin(b = 10)),
    plot.subtitle = element_text(hjust = 0.5, margin = margin(b = 15)),
    legend.position = "none",
    axis.line = element_line(color = "gray80"),
    axis.text = element_text(color = "gray30"),
    axis.ticks = element_line(color = "gray80"),
    axis.title.x = element_text(margin = margin(t = 10)),
    plot.caption = element_text(margin = margin(t = 15), hjust = 0),
    plot.margin = margin(20, 20, 30, 20)
  ) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray70") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray30")

@@

# Save the plot
ggsave(
  filename = file.path(reanalysis_data_dir, "forest_plot_ps_methods_imputed.png"),
  plot = forest_plot,
  width = 10,
  height = 8,
  dpi = 300
)

@@

## Summary Table

```{r summary-table}
# Display the summary table
kable(summary_table, 
      caption = "Pooled Effect Estimates Across Propensity Score Methods",
      col.names = c("Method", "Odds Ratio", "95% CI", "P-value", "N Imputations"),
      align = c("l", "r", "c", "r", "r"))

# Save to CSV for external use
write.csv(summary_table, 
          file.path(reanalysis_data_dir, "ps_methods_comparison_table.csv"),
          row.names = FALSE)

log_message("Summary table saved to CSV")
```

## Method Comparison

```{r method-comparison}
# Compare all methods to twang_gbm as reference
print_method_comparison(pooled_results, reference_method = "twang_gbm")
```

## Individual Method Details

```{r method-details}
# Show detailed results for each method
cat("\n", paste(rep("=", 60), collapse = ""), "\n")
cat("DETAILED RESULTS BY METHOD\n")
cat(paste(rep("=", 60), collapse = ""), "\n\n")

for (method_name in names(pooled_results)) {
  res <- pooled_results[[method_name]]
  if (res$success) {
    cat(paste0("Method: ", method_name, "\n"))
    cat(paste0("  Odds Ratio: ", sprintf("%.3f", res$or), "\n"))
    cat(paste0("  95% CI: (", sprintf("%.3f", res$or_ci_lower), 
              ", ", sprintf("%.3f", res$or_ci_upper), ")\n"))
    cat(paste0("  P-value: ", sprintf("%.4f", res$p_value), "\n"))
    cat(paste0("  Log OR (SE): ", sprintf("%.3f", res$estimate), 
              " (", sprintf("%.3f", res$se), ")\n"))
    cat(paste0("  N imputations: ", res$n_imputations, "\n"))
    cat(paste0("  Degrees of freedom: ", sprintf("%.1f", res$df), "\n\n"))
  }
}
```

## Diagnostics Summary

```{r diagnostics-summary}
# Summarize success rates for each method
cat("\n", paste(rep("=", 60), collapse = ""), "\n")
cat("PROPENSITY SCORE ESTIMATION SUCCESS RATES\n")
cat(paste(rep("=", 60), collapse = ""), "\n\n")

for (method_name in names(ps_results)) {
  method_results <- ps_results[[method_name]]
  n_success <- sum(sapply(method_results, function(x) !is.null(x) && x$success))
  n_total <- length(method_results)
  success_rate <- 100 * n_success / n_total

  cat(sprintf("%-20s: %d/%d successful (%.1f%%)\n", 
             method_name, n_success, n_total, success_rate))
}
```

@@

This analysis compared **`r length(pooled_results)` different propensity score methods** across **`r config$n_imputations` multiply imputed datasets**. Results were pooled using Rubin's rules to properly account for between-imputation variability.

### Key Findings:

```{r key-findings}
# Find min and max effect sizes
or_values <- sapply(pooled_results, function(x) if(x$success) x$or else NA)
or_values <- or_values[!is.na(or_values)]

min_method <- names(which.min(or_values))
max_method <- names(which.max(or_values))

cat(paste0("- Method with smallest effect: ", min_method, 
          " (OR = ", sprintf("%.3f", min(or_values)), ")\n"))
cat(paste0("- Method with largest effect: ", max_method, 
          " (OR = ", sprintf("%.3f", max(or_values)), ")\n"))
cat(paste0("- Range of odds ratios: ", sprintf("%.3f to %.3f", 
          min(or_values), max(or_values)), "\n"))

if ("twang_gbm" %in% names(or_values)) {
  twang_or <- or_values["twang_gbm"]
  cat(paste0("- twang GBM (primary method): OR = ", sprintf("%.3f", twang_or), "\n"))

  # How many methods are within 10% of twang
  within_10pct <- sum(abs(or_values / twang_or - 1) <= 0.10)
  cat(paste0("- Methods within 10% of twang GBM: ", within_10pct, 
            " out of ", length(or_values), "\n"))
}
```

### Analysis Details:
- **Estimand:** ATT (Average Treatment Effect in the Treated)
- **Imputation method:** Multiple imputation using mice
- **Pooling method:** Rubin's rules
- **Outcome model:** Survey-weighted logistic regression with cluster-robust SEs

## Session Info

```{r session-info}
# Finalize logging
log_message("Analysis completed successfully")
finalize_log(success = TRUE)

# Print session information
sessionInfo()
```
